Open archaeology, open source?

### 2023-10-12 analysis/_01-introduction.qmd
Mostly writing uodates.
Updated/corrected bibliographic references. Updated certain parts of the text for clarity. Some other scattered notes and superficial changes.

However, academic open source has a complicated relationship with open source as practiced by professional software developers, which has its own distinct history and is framed by different objectives, challenges, and value regimes. Despite this, the open science movement, within which open archaeology emerged, draws direct inspiration from open source. For instance, the Open Knowledge Foundation [-@openknowledgefoundation2023] publishes a widely accepted definition of "open" in the context of scholarly communication that explicitly refers to the definition of "open source" published by the Open Source Initiative [-@opensourceinitiative2007], an authoritative open source advocacy group. The open science movement further mimics open source by operationalizing scholarly communication through technical infrastructures and protocols that closely resemble systems and processes designed to develop open source software (e.g., the use of plain text, line-resolution version control, emphasis on formal licencing, the general hacker aesthetic). However, academic work, including the development of academic software, differs significantly from the work involved in massive open source projects that literally run the internet, such as the Linux kernel, openSSL, and the Firefox web browser. While they may use similar tools and technical protocols to manage coding operations, the open science and open source movements are governed by different social and professional warrants and interests. In other words, publishing code openly on the web has different meanings, impacts and implications for archaeologists and professional software developers [@kelty2008 xx].

@@

<!-- Might be getting a bit ahead of ourselves here. --JR -->That being said, archaeologists are prolific software developers. This is made evident by open-archaeo, a comprehensive list of open source archaeological software and resources that indexes XXX records as of the time of publication. But beyond simply making their code available on the web, do archaeologists also implement social strategies to advance open source ideals? Does archaeological open source actually help achieve greater transparency, sustainability, and community participation? And if not, what does it actually achieve?

@@

(1) we identify *what* kinds of software archaeologists are making;
(2) and evaluate *how* archaeologists create these tools, with particular emphasis on practices of collaboration.

>>>>

However, academic open source has a complicated relationship with open source as practiced by professional software developers, which has its own distinct history and is framed by different objectives, challenges, and value regimes. Despite this, the open science movement, within which open archaeology emerged, draws direct inspiration from open source. For instance, the Open Knowledge Foundation [-@openknowledgefoundation2015] publishes a widely accepted definition of "open" in the context of scholarly communication that explicitly refers to the definition of "open source" published by the Open Source Initiative [-@opensourceinitiative2007], an authoritative open source advocacy group. The open science movement further mimics open source by operationalizing scholarly communication through technical infrastructures and protocols that closely resemble systems and processes designed to develop open source software (e.g., the use of plain text, line-resolution version control, emphasis on formal licencing, the general hacker aesthetic). However, academic work, including the development of academic software, differs significantly from the work involved in massive open source projects that literally run the internet, such as the Linux kernel, openSSL, and the Firefox web browser. While they may use similar tools and technical protocols to manage coding operations, the open science and open source movements are governed by different social and professional warrants and interests. In other words, publishing code openly on the web has different meanings, impacts and implications for archaeologists and professional software developers [@ratto2007; @kelty2008 chapter 9].

@@

<!-- Might be getting a bit ahead of ourselves here. --JR -->That being said, archaeologists are prolific software developers. This is made evident by open-archaeo, a comprehensive list of open source archaeological software and resources that indexes <!-- TODO: encode -->XXX records, as of the time of this article's publication. But beyond simply making their code available on the web, do archaeologists also implement social strategies to advance open source ideals? Does archaeological open source actually help achieve greater transparency, sustainability, and community participation? And if not, what does it actually achieve?

@@

(1) we identify *what* kinds of software archaeologists are making; and
(2) we evaluate *how* archaeologists create these tools, with particular emphasis on practices of collaboration.

### 2023-10-12 analysis/_02-background.qmd
Mostly writing uodates.
Updated/corrected bibliographic references. Updated certain parts of the text for clarity. Some other scattered notes and superficial changes.

Open source is a software development model that prioritizes transparent work processes. Initially driven by the idea that computer users should be free to understand and manipulate the software that they install on their computers (e.g., "free software", as initially conceived by the Free Software Foundation), open source has become a means of collaborative software development [@kelty2008: chapter 3, especially starting at page 99]. By putting one's code on the web without restriction on how it may be used or manipulated, this encourages creativity to flourish as people contribute to help improve the code base. Software thus emerges from the coordinated labour of worldwide volunteers, who shape the product according to the collective vision [@kelty2008 on the emacs saga]. An open code base may also be used to support alternative projects whose missions diverge from the original plan, and an entire project may be "forked", or taken in a new direction if contributors are dissatisfied with how core developers run things.

Open source has traditionally been referred to as being based on meritocratic principles. A good test of whether a contribution should be included in a published software release is whether it is functional [@kelty2008 220]. Moreover, with more eyes looking over a code base, it is easier to identify flaws with a contribution and flag potential bugs or security issues. This is all done in the spirit of producing functional code, and in ideal circumstances faulty contributions will be corrected before inclusion. Personal ego is minimized in favour of co-creating stable and functional outcomes [@coleman2012; @oneil2009 on the Debian saga].

However, this is not the same as saying that open source is completely anarchic or based on the "wisdom of crowds". In fact, successful open source projects incorporate complex organizational structures, governance strategies, and forms or social mediation to help delegate and vet contributions made by distributed participants [@oneil2009]. They rely on, rather than eschew, institutional support structures, in order to motivate work, keep volunteer maintainers involved, and generally ensure that the project can be sustained over the long-term. Open source is more than just putting your code online; to be successful, it requires participation in a social experience [@kelty2008, Ratto dissertation and derived works].

@@

The open science movement comprises a series of practices and principles intended to make research more accessible, transparent, and efficient. Although the concept of "open" is somewhat nebulous, in terms of its abstract definition and with regards to what real-world applications count as being open, one commonly-cited definition describes content that "can be freely used, modified, and shared by anyone for any purpose" [@openknowledgefoundation2023]. This definition does not state what open is for, how to be open, or any sort of social or discursive framing behind the open movement. However, most open science advocates (including archaeologists, as elicited by @beck2012 and @marwick2017) claim that they are motivated by a desire to facilitate novel research opportunities, make participation in scientific research more equitable, reclaim science as a public good, and enhance how findings are validated and legitimized.

@@

The adoption of open source development models among archaeologists is generally informed by the broader open science movement, which is motivated by a genuine desire to facilitate novel research opportunities, to make participation in scientific research more equitable, to reclaim science as a public good, and to enhance the means of validating findings. However, the predominant concern with implementing best tools to use, adopting optimal data processing pipelines, and tying into global, web-based infrastructures, protocols and standards [cf. @kansa2014; @kintigh2015; @roosevelt2015] distract from fundamental tensions and contradictions regarding the actual value of working in the open. For instance, Faniel et al. [-@faniel2013 299-301], Atici et al. [-@atici2013 676-677], @huggett2018, @sobotkova2018, and @opitz2021 demonstrate that to make the reuse of archaeological data feasible and useful in a practical sense, it is necessary to re-introduce social friction that these infrastructures are designed to eliminate. In other words, the pressures and circumstances of being an archaeologist and doing archaeological research assert themselves when attempting to make practical use of these infrastructures, and therefore must be accounted for in their design and implementation. Here we aim to identify similar sources of dissonance with regards to the promise, potential, and actual implementation of open source software development models among archaeologists.

>>>>

Open source is a software development model that prioritizes transparent work processes. Initially driven by the idea that computer users should be free to understand and manipulate the software that they install on their computers (e.g., "free software", as initially conceived by the Free Software Foundation), open source has become a means of collaborative software development [@kelty2008: chapter 3, especially starting at page 99]. By putting one's code on the web without restriction on how it may be used or manipulated, this encourages creativity to flourish as people contribute to help improve the code base. Software thus emerges from the coordinated labour of worldwide volunteers, who shape the product according to the collective vision [see @kelty2008 on the emacs saga for a quintessential example]. An open code base may also be used to support alternative projects whose missions diverge from the original plan, and an entire project may be "forked", or taken in a new direction if contributors are dissatisfied with how core developers run things.

Open source has traditionally been referred to as being based on meritocratic principles [@raymond1999 39]. A good test of whether a contribution should be included in a published software release is whether it is functional [@kelty2008 220]. Moreover, with more eyes looking over a code base, it is easier to identify flaws with a contribution and flag potential bugs or security issues [@raymond1999 27-30]. This is all done in the spirit of producing functional code, and in ideal circumstances faulty contributions will be corrected before inclusion. Personal ego is minimized in favour of co-creating stable and functional outcomes [@raymond1999 39-41].

However, this is not the same as saying that open source is completely anarchic or based on the "wisdom of crowds". In fact, successful open source projects incorporate complex organizational structures, governance strategies, and forms or social mediation to help delegate and vet contributions made by distributed participants [@oneil2009]. They rely on, rather than eschew, institutional support structures, in order to motivate work, keep volunteer maintainers involved, and generally ensure that the project can be sustained over the long-term. Open source is more than just putting your code online; to be successful, it requires participation in a social experience [@ratto2003; @kelty2008].

@@

The open science movement comprises a series of practices and principles intended to make research more accessible, transparent, and efficient. Although the concept of "open" is somewhat nebulous, in terms of its abstract definition and with regards to what real-world applications count as being open, one commonly-cited definition describes content that "can be freely used, modified, and shared by anyone for any purpose" [@openknowledgefoundation2015]. This definition does not state what open is for, how to be open, or any sort of social or discursive framing behind the open movement. However, most open science advocates (including archaeologists, as elicited by @beck2012 and @marwick2017) claim that they are motivated by a desire to facilitate novel research opportunities, make participation in scientific research more equitable, reclaim science as a public good, and enhance how findings are validated and legitimized.

@@

The adoption of open source development models among archaeologists is generally informed by the broader open science movement, which is motivated by a genuine desire to facilitate novel research opportunities, to make participation in scientific research more equitable, to reclaim science as a public good, and to enhance the means of validating findings. However, the predominant concern with implementing best tools to use, adopting optimal data processing pipelines, and tying into global, web-based infrastructures, protocols and standards [cf. @kansa2014; @kintigh2015; @roosevelt2015] distract from fundamental tensions and contradictions regarding the actual value of working in the open. For instance, Faniel et al. [-@faniel2013 299-301], Atici et al. [-@atici2013 676-677], Huggett [-@huggett2018; -@huggett2022], @sobotkova2018, @opitz2021, @haciguzeller2021, and @batist2023a demonstrate that to make the reuse of archaeological data feasible and useful in a practical sense, it is necessary to re-introduce social friction that these infrastructures are designed to eliminate. In other words, the pressures and circumstances of being an archaeologist and doing archaeological research assert themselves when attempting to make practical use of these infrastructures, and therefore must be accounted for in their design and implementation. Here we aim to identify similar sources of dissonance with regards to the promise, potential, and actual implementation of open source software development models among archaeologists.

### 2023-10-12 analysis/_03-data_methodology.qmd
Mostly writing uodates.
Updated/corrected bibliographic references. Updated certain parts of the text for clarity. Some other scattered notes and superficial changes.

Open-archaeo is a relatively comprehensive list.
While our initial intention was to only list open source software, its scope has expanded to include all software created by and for archaeologists.
Apart from regular updates by its primary maintainer (ZB), it has been expanded by a wider network of contributors and has benefited from the wider range of domain specialisms this has brought.
However, there are a number of areas that, for practical reasons, are not well-covered.
<!-- TODO: next three sentences are a bit repeatative, can we more concisely state what OA doesn't cover well? -->
Code written before archaeologists started using collaborative software development platforms such as GitHub is generally missing.
It also lacks references to code that is not available on the web.
Open-archaeo is also limited by the experiences of its primary maintainers, and we welcome anyone, especially domain specialists who are familiar with the kinds of tools commonly used in their specific fields, to contribute.

@@

We augmented the open-archaeo's list of with projects with information on its activity on GitHub, where applicable.
This included metadata on the GitHub repository and the programming languages it uses, as well as detailed data on each repository's individual contributors, commits, issues and comments (see @tbl-glossary for a glossary of git/GitHub terminology used in this paper).
This data was collected via GitHub's public API using the R package `gh` [@csardi2023].
We opted to only collect repository data from GitHub because it is the most popular forge platform used by open-archaeo projects (@tbl-forges).
This means that projects that don't use version control (`r percent(p_no_vc)` of the total), or host it elsewhere (`r percent(p_no_github)` of the total), are excluded from these parts of the analysis, though we were still able to perform an analysis of their contents and authorship from the data included in open-archaeo itself.

@@

The data and analysis code are available in our research compendium on GitHub, which is also indexed on Zenodo.

>>>>

Open-archaeo is a relatively comprehensive list.
While our initial intention was to only list open source software, its scope has expanded to include all software created by and for archaeologists.
Apart from regular updates by its primary maintainer (ZB), it has been expanded by a wider network of contributors and has benefited from the wider range of domain specialisms this has brought.
However, open-archaeo generally lacks software written before archaeologists started using collaborative software development platforms such as GitHub, and software that is not shared on the web at all.
The dataset is also limited by the experiences of its primary maintainers.
We welcome anyone, especially domain specialists who are familiar with the kinds of tools commonly used in their specific fields, to help fill in these gaps.

@@

Where applicable, we obtained more detailed information about each repository's contents and contribution histories from the GitHub API using the R package `gh` [@csardi2023].
This included metadata on the GitHub repository and the programming languages it uses, as well as detailed data on each repository's individual contributors, commits, issues and comments (see @tbl-glossary for a glossary of git/GitHub terminology used in this paper).
We opted to only collect repository data from GitHub because it is the most popular forge platform used by open-archaeo projects (@tbl-forges).
This means that projects that do not use version control (`r percent(p_no_vc)` of the total), or host it elsewhere (`r percent(p_no_github)` of the total), are excluded from these parts of the analysis, though we were still able to perform an analysis of their contents and authorship from the data included in open-archaeo itself.

@@

The data and analysis code are available in our research compendium on GitHub, which is also indexed on Zenodo. <!-- TODO: add hyperlinks to these resources -->

### 2023-10-12 analysis/_04-open_archaeology.qmd
Mostly writing uodates.
Updated/corrected bibliographic references. Updated certain parts of the text for clarity. Some other scattered notes and superficial changes.

`r percent(p_platform_r)` of projects are extensions to the statistical programming language R, making it the most widely-used 'platform' by a large margin (@tbl-platforms).
Python, another programming language, is also relatively popular <!-- TODO: percent -->, as are plugins for the open source geographic information system QGIS <!-- TODO: percent -->.
Beyond that, there is a rather fragmented landscape of plugins for other desktop software (e.g. AutoCAD, ArcGIS), a number of lesser used programming languages, and a genre consisting of custom forms and spreadsheet templates.
Many of these are targeted by only one or two developers; the larger platforms tend to be more diverse.

@@

We also annotated each record with 'tags' that describe aspects of archaeological work that each tool contributes to [@fig-tags].
The most common tags unsurprisingly deal with work that naturally benefits from advanced information processing afforded by computers, such as statistical analysis, sample calibration, geographical analysis, data management, and chronological modelling.
Educational resources and practical guides are also well represented due to the web's usefulness as a medium for sharing and communication.

>>>>

`r percent(p_platform_r)` of projects are extensions to the statistical programming language R, making it the most widely-used platform by a large margin (@tbl-platforms).
Python, another programming language, is also relatively popular <!-- TODO: percent -->, as are plugins for the open source geographic information system QGIS <!-- TODO: percent -->.
Beyond that, there is a rather fragmented landscape of plugins for other desktop software (e.g. AutoCAD, ArcGIS), a number of lesser used programming languages, and a genre consisting of custom forms and spreadsheet templates.
Many of these are targeted by only one or two developers; the larger platforms tend to be more diverse.

@@

We also annotated each record with "tags" that describe aspects of archaeological work that each tool contributes to [@fig-tags].
The most common tags unsurprisingly deal with work that naturally benefits from advanced information processing afforded by computers, such as statistical analysis, sample calibration, geographical analysis, data management, and chronological modelling.
Educational resources and practical guides are also well represented due to the web's usefulness as a medium for sharing and communication.

### 2023-10-12 analysis/_05-collaboration.qmd
Mostly writing uodates.
Updated/corrected bibliographic references. Updated certain parts of the text for clarity. Some other scattered notes and superficial changes.

```{r fig-issues-comments}
#| fig-cap: Use of issues and comments
# TODO: include zeroes
# TODO: why are there no twos?
# TODO: throw stars in here too?
fig_issues_hist <- oarch_gh %>%
  unnest(gh_issues) %>%
  drop_na(title) %>%
  group_by(repo) %>%
  summarise(n = n(), .groups = "drop") %>%
  ggplot(aes(x = n)) +
  geom_histogram() +
  scale_x_log10() +
  labs(x = "Number of issues", y = "Repositories") +
  theme_oarch()

@@

```{r plot-gh_contributors-histogram}
oarch %>%
  drop_na(gh_repo) %>%
  unnest(gh_contribs) %>%
  group_by(gh_repo) %>%
  summarise(n = n(), .groups = "drop") %>%
  ggplot(aes(x = n)) +
  geom_histogram() +
  scale_x_log10() +
  labs(
    title = "Repository contributors",
    x = "Number of contributors", y = "Repositories",
    caption = "Data: open-archaeo.info and GitHub API"
  ) +
  theme_oarch()
```

* How many people are involved?
  *  Conversely, how many projects are individual people involved in?

Perhaps unfortunately then, we also find that nearly 60% of projects have only a single contributor, and projects with more than a dozen are very rare indeed \[\<3%\]. A "contribution", in this case, is a commit, issue, or comment applied by a user to a repository.
<!-- TODO: expand -->

```{r plot-contrib-dist}
#TODO: Re-calculate these numbers to ensure they are up to date. Translate the first point to be about how many repos have a quarter of contributions by a single person, so that it corresponds with the vertical lines on the graphic, which are now divided into quarters rather than thirds.
oarch %>%
  unnest(gh_contribs) %>%
  drop_na(contributor) %>%
  group_by(item_name) %>%
  mutate(n_contributors = n()) %>%
  filter(n_contributors > 1) %>%
  mutate(big_boy = ifelse(contributions == max(contributions),
                          "Leading contributor", "Other contributors")) %>% 
  group_by(item_name, big_boy) %>%
  summarise(contributions = sum(contributions)) %>%
  mutate(
    p_contributions = contributions / sum(contributions),
    p_big_boy = first(p_contributions)
  ) %>%
  ungroup() %>%
  mutate(item_name = fct_reorder(item_name, p_big_boy)) %>% 
  ggplot(aes(x = item_name, y = contributions,
             fill = fct_relevel(big_boy, c("Other contributors",
                                           "Leading contributor")))) +
  geom_col(position = "fill", colour = "white", size = 0.1) +
  geom_hline(yintercept = c(0.25, 0.5, 0.75), alpha = 0.5) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_highcontrast(guide = guide_legend(reverse = TRUE)) +
  labs(title = "Repository contributors by proportion of contributions",
       x = NULL, y = "Contributions",
       fill = NULL,
       caption = "Data: open-archaeo.info and GitHub API") +
  coord_flip() +
  theme_oarch() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )
```

Furthermore, when we break down the contributions made by all contributors in multi-contributor projects, we also see that work tends to be distributed very unevenly. Of multi-contributor repositories, 92% have a third of contributions made by a single person, 65% have half of all contributions made by a single person, and 36% have three quarters of all contributions made a single person.

@@

The prototypical collaborative open source project on GitHub has a core group of developers (often a single maintainer) that regularly commit new code, a wider network of collaborators that contribute through pull requests, and an active user base that create and comment on issues, who have indicated their support for the project by starring its repository.
The data we collected on open-archaeo repositories on GitHub show that uptake of its collaboration features is highly uneven, with very few projects resembling the open source ideal.
The majority of projects are short-lived, with few contributors and a small number of commits.
Only a fraction of repositories use GitHub's extended collaboration features (pull requests, issues, comments) and those that do use them in a limited way.
<!-- TODO: note that we don't know if maintainers actually aimed for that model, or leave that for the discussion? -->
Conversely, a small number of projects (<!-- TODO: do these share any characteristics in terms of platform or subject? -->) do appear to fit the aspirational open source model, with a large number of contributors maintaing a long-lived and highly-active repository.
Archaeologists active on GitHub also make markedly <!-- TODO: verify --> greater use of GitHub's social media-like features (stars, following) than more direct collaborative actions.

>>>>

```{r fig-issues-comments}
#| fig-cap: Use of issues and comments.
# TODO: include zeroes
# TODO: why are there no twos?
# TODO: throw stars in here too?
fig_issues_hist <- oarch_gh %>%
  unnest(gh_issues) %>%
  drop_na(title) %>%
  group_by(repo) %>%
  summarise(n = n(), .groups = "drop") %>%
  ggplot(aes(x = n)) +
  geom_histogram() +
  scale_x_log10() +
  labs(x = "Number of issues", y = "Repositories") +
  theme_oarch()

@@

```{r plot-gh_contributors-histogram}
#| fig-cap: Number of contributors.
oarch %>%
  drop_na(gh_repo) %>%
  unnest(gh_contribs) %>%
  group_by(gh_repo) %>%
  summarise(n = n(), .groups = "drop") %>%
  ggplot(aes(x = n)) +
  geom_histogram() +
  scale_x_log10() +
  labs(
    title = "Repository contributors",
    x = "Number of contributors", y = "Repositories",
    caption = "Data: open-archaeo.info and GitHub API"
  ) +
  theme_oarch()
```

Perhaps unfortunately then, we also find that nearly 60% of projects have only a single contributor, and projects with more than a dozen are very rare indeed \[\<3%\] (@plot-gh_contributors-histogram). A "contribution", in this case, is a commit, issue, or comment applied by a user to a repository.
<!-- TODO: expand -->

```{r plot-contrib-dist}
#| fig-cap: Distribution of contributions in multi-contributor projects.
#TODO: Re-calculate these numbers to ensure they are up to date. Translate the first point to be about how many repos have a quarter of contributions by a single person, so that it corresponds with the vertical lines on the graphic, which are now divided into quarters rather than thirds.
oarch %>%
  unnest(gh_contribs) %>%
  drop_na(contributor) %>%
  group_by(item_name) %>%
  mutate(n_contributors = n()) %>%
  filter(n_contributors > 1) %>%
  mutate(big_boy = ifelse(contributions == max(contributions),
                          "Leading contributor", "Other contributors")) %>% 
  group_by(item_name, big_boy) %>%
  summarise(contributions = sum(contributions)) %>%
  mutate(
    p_contributions = contributions / sum(contributions),
    p_big_boy = first(p_contributions)
  ) %>%
  ungroup() %>%
  mutate(item_name = fct_reorder(item_name, p_big_boy)) %>% 
  ggplot(aes(x = item_name, y = contributions,
             fill = fct_relevel(big_boy, c("Other contributors",
                                           "Leading contributor")))) +
  geom_col(position = "fill", colour = "white", size = 0.1) +
  geom_hline(yintercept = c(0.25, 0.5, 0.75), alpha = 0.5) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_highcontrast(guide = guide_legend(reverse = TRUE)) +
  labs(title = "Repository contributors by proportion of contributions",
       x = NULL, y = "Contributions",
       fill = NULL,
       caption = "Data: open-archaeo.info and GitHub API") +
  coord_flip() +
  theme_oarch() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )
```

Furthermore, when we break down the contributions made by all contributors in multi-contributor projects, we also see that work tends to be distributed very unevenly (@plot-contrib-dist). Of multi-contributor repositories, 92% have a third of contributions made by a single person, 65% have half of all contributions made by a single person, and 36% have three quarters of all contributions made a single person.

@@

The prototypical collaborative open source project on GitHub has a core group of developers (often a single maintainer) that regularly commit new code, a wider network of collaborators that contribute through pull requests, and an active user base that create and comment on issues, who have indicated their support for the project by starring its repository.
The data we collected on open-archaeo repositories on GitHub show that uptake of its collaboration features is highly uneven, with very few projects resembling the open source ideal.
The majority of projects are short-lived, with few contributors and a small number of commits.
Only a fraction of repositories use GitHub's extended collaboration features (pull requests, issues, comments) and those that do use them in a limited way.
<!-- TODO: note that we don't know if maintainers actually aimed for that model, or leave that for the discussion? -->
Conversely, a small number of projects (<!-- TODO: do these share any characteristics in terms of platform or subject? -->) do appear to fit the aspirational open source model, with a large number of contributors maintaining a long-lived and highly-active repository.
Archaeologists active on GitHub also make markedly <!-- TODO: verify --> greater use of GitHub's social media-like features (stars, following) than more direct collaborative actions.

### 2023-10-12 analysis/openarchaeo_collaboration.qmd
Mostly writing uodates.
Updated/corrected bibliographic references. Updated certain parts of the text for clarity. Some other scattered notes and superficial changes.

  In this paper, we investigate modes of collaboration in this emerging community of practice using open-archaeo (https://open-archaeo.info/), a curated list of 500+ active open projects in archaeology, augmented where available with data on their activity on GitHub. We apply network analysis to map the networks of collaboration that have emerged among archaeologists using GitHub, and conduct a quantitative exploratory analysis to characterize the nature and intensity of these collaborations. Our results show an uneven adoption of open source collaborative practices beyond the basic use of `git` as a version control system and GitHub to host source code. A majority of repositories have 1–3 contributors, with only a few projects distinguished by an active and diverse developer base. More community-oriented features, such as GitHub's 'issues' and 'comments', are used in only a minority of repositories. Preliminary findings from ongoing work also shows how collaborative behaviours vary across different subcommunities. By analyzing interactions among archaeologists using the GitHub platform, including the ways they contribute to collective code bases, we demonstrate how archaeological software engineering is beginning to foster new kinds of collaborative commitments that blend open source protocols, DIY work ethic, conventional forms of scholarly communication, and cultural practices unique to archaeology.
bibliography: references.bib
format:
  html:
    page-layout: article
    toc: true
    number-sections: true
    embed-resources: true
---

>>>>

  In this paper, we investigate modes of collaboration in this emerging community of practice using open-archaeo [open-archaeo.info](open-archaeo.info), a curated list of 500+ active open projects in archaeology, augmented where available with data on their activity on GitHub. We apply network analysis to map the networks of collaboration that have emerged among archaeologists using GitHub, and conduct a quantitative exploratory analysis to characterize the nature and intensity of these collaborations. Our results show an uneven adoption of open source collaborative practices beyond the basic use of `git` as a version control system and GitHub to host source code. A majority of repositories have 1–3 contributors, with only a few projects distinguished by an active and diverse developer base. More community-oriented features, such as GitHub's 'issues' and 'comments', are used in only a minority of repositories. Preliminary findings from ongoing work also shows how collaborative behaviours vary across different subcommunities. By analyzing interactions among archaeologists using the GitHub platform, including the ways they contribute to collective code bases, we demonstrate how archaeological software engineering is beginning to foster new kinds of collaborative commitments that blend open source protocols, DIY work ethic, conventional forms of scholarly communication, and cultural practices unique to archaeology.
bibliography: references.bib
format:
  html:
    page-layout: article
    toc: true
    number-sections: true
    embed-resources: true
---
